Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-07 doc/thesis/methods_results.tex
More exposition for ABC; rewrote discussion of real data

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. A
sudden precipitous drop in \gls{ESS} would indicate that only a small number of
particles had non-zero importance weights, which would result in a very poor
Monte Carlo approximation to the target distribution. This situation is
referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
confused with the \gls{BA} model parameter)}} controls the decay rate.
{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive}, \uline{the
parameter is called $\alpha$, but to avoid confusion with the \gls{BA}
parameter of the same name, we will refer to it here as $\alpha_{\ESS}$.}}
The tolerance $\varepsilon_i$ is chosen to satisfy
\[
  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
\]
where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
proposal distributions
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
\]
where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
simulated according to $\theta'$. In our implementation, $q$ is either a
Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
parameters. For the Poisson proposals, the number of steps to move the particle
is drawn from a Poisson distribution, and the direction in which to move the
particle is chosen uniformly at random. For both proposals, the variance was
set equal to twice the empirical variance of the particles,
following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
kernels are
\[
  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
\]
When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
with the weight update 
\begin{align*}
  w_i(x) 
    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
\end{align*}
In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
simulated datasets that are still inside the new threshold. 
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel, and the algorithm will be stopped when either of
these termination conditions is reached. The latter condition stops the
algorithm when the particles are not moving around very much, implying little
change in the estimated target.

@@

\uline{We have shown that both the normalizing constant $\Pr(T)$ and the
likelihood $\Pr(T \mid \theta)$ are likely computationally prohibitive to
calculate. If this is the case, the problem of fitting contact network models
to phylogenies seems to be of the \emph{doubly-intractible}
type~\autocite{murray2012mcmc}, which would imply that these models are not
amenable to neither \gls{ML} nor Bayesian inference techniques. Although both
methods are able to cope with an intractible normalizing constant (for example,
by local search for \gls{ML} or Bayesian \gls{MCMC}), neither can avoid the
intractable likelihood calculations. This justifies the use of \gls{ABC}, which
is a likelihood-free method.}

@@

@@

{\color{blue}\uline{
Our first computational experiment was designed as an exploratory analysis of
the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
\gls{m}, and \gls{N}. The objective of this experiment was to determine whether
any of the four parameters were identifiable from the shape of the transmission
tree, as quantified by the tree kernel. Each of the \gls{BA} model parameters
was varied one at a time, while holding the other parameters at fixed, known
values. Contact networks were generated according to each set of parameter
values, and transmission trees were simulated over the networks. We then
evaluated how well a classifier based on the tree kernel could differentiate
the trees simulated under distinct parameter values. If the classifier's
cross-validation accuracy was high, this could be taken as an indication that
the parameter in question was identifiable in the range of values considered. 
A caveat of this preliminary analysis is that, since all parameters but one
were held at known values, nothing could be said about the identifiability of
\emph{combinations} of parameters; this issue will be explored later by jointly
estimating all parameters with \gls{ABC}} (\cref{subsec:abc}).

\uline{In addition to testing for identifiability, a secondary objective of
this analysis was to validate the use of the tree kernel as a distance measure
for \gls{ABC} in our context. As discussed in} \cref{sec:abc}, \uline{the
choice of distance function is extremely important for the accuracy of the
\gls{ABC} approximation to the posterior. Therefore, we evaluated two
additional tree statistics in the same manner as we evaluated the tree kernel
(that is, by constructing and testing a classifier). First, we considered
Sackin's index~\autocite{shao1990tree}, which measures the degree of imbalance
or asymmetry in a phylogeny} (see \cref{subsec:treeshape}). \uline{Sackin's
index is widely used for characterizing
phylogenies~\autocite{frost2013modelling} and has been demonstrated to vary
between transmission trees simulated under different contact network
types~\autocite{leventhal2012inferring}. Sackin's index does not take branch
lengths into account, considering only the tree's topology. The other statistic
we considered was the \gls{nltt}~\autocite{janzen2015approximate}, which
compares two trees based on normalized distributions of their branching times}
(see \cref{sec:treeshape}).  \uline{In contrast with Sackin's index, the
\gls{nltt} does not explicity consider the trees' topologies, but it does use
their normalized branch lengths. While the \gls{nltt} is a newly developed
statistic not yet in widespread use, the unnormalized
\gls{ltt}~\autocite{nee1992tempo} was the basis of seminal early work
extracting epidemiological information from
phylodynamics~\autocite{holmes1995revealing}. We expected the tree kernel to
classify the \gls{BA} parameters more accurately than either Sackin's index or
the \gls{nltt}, since the tree kernel takes both topology and branch lengths
into account. }}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{gridsearch-example}
  \caption[Grid search estimates of \gls{BA} model parameters.]{Point estimates
      (center lines) and 75\% (boxes) and 95\% (whiskers) highest density
      intervals for each \gls{BA} model parameter, obtained using grid search.
      Networks and transmission trees were simulated over a grid of values for
       each parameter while holding the others fixed. For a subset of the grid
      values ($x$-axis), test networks and trees were created and compared to
      each tree on the grid using the tree kernel. The kernel scores along the
      grid were normalized to resemble a probability distribution, from which
      the mode and highest density interval were calculated. Shown values
      correspond to one replicate experiment, with trees of size 500.
  } 
  \label{fig:gridest}
\end{figure}

@@

{\color{blue}
\uline{Across all simulations, the median [\gls{IQR}] absolute errors of the parameter 
estimates obtained with \software{netabc} were
    0.11 
    [0.03 - 
    0.25]
for \gls{alpha},
    492 
    [294 - 
    782]
for \gls{I},
    1 
    [0 - 
    1]
for \gls{m}, and
    4153 
    [3660 - 
    4489]
for \gls{N}. These errors comprised, respectively,
    6\%,
    11\%,
    17\%, and
    29\%
of the regions of nonzero prior density. Average \gls{HPD} interval widths were
    0.68,
    2454,
    3.01, and
    12046,
representing 
    34\%,
    55\%,
    50\%, and
    83\%
of the nonzero prior density regions. Point estimates of \gls{I} were upwardly
biased: \gls{I} was overestimated in
    69
out of 
    72
simulations
    (4\%).
The estimates for \gls{m} and \gls{N} were similar across all simulations
(median [IQR] point estimates
    3
    [3 - 
    3]
and 
    9153
    [8660 - 
    9489])
regardless of the true values of any of the \gls{BA} parameters.}

@@

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Thu Jun  2 15:02:59 2016
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
Parameter & Estimate & Standard error & $p$-value \\ 
  \hline
(Intercept) & $2$ & $0.6$ & $0.01$ \\ 
  $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
  I & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
  m & $0.5$ & $0.2$ & $0.01$ \\ 
   \hline
\end{tabular}
\caption[Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters. GLM was fitted with a Gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glmalpha}
\end{table}
% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Thu Jun  2 15:02:59 2016
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
Parameter & Estimate & Standard error & $p$-value \\ 
  \hline
(Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
  $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
  I & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
  m & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
   \hline
\end{tabular}
\caption[Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters. GLM was fitted with a gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glmI}
\end{table}

@@

Empirical studies of contact networks often report the exponent $\gamma$ of the
power law degree distribution. To compare our results to the literature, we
simulated 100 networks each according to the \gls{MAP} parameter estimates
obtained for each investigated dataset. The power-law exponent $\gamma$ was
calculated for each network using the \software{fit\_power\_law} function in
\software{igraph}, with the `R.mle' implementation. The median of the 100
$\gamma$ values was taken as a point estimate for the associated dataset.

@@

@@

We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
and found substantial heterogeneity among the parameter estimates
(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
for each dataset are shown in
\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
estimated $\alpha$ values near unity for the prior allowing $m = 1$ (\gls{MAP}
estimates [95\% \gls{HPD}] 
  0.76 
  [0.07 - 
   1.34]
and
  0.61 
  [0.01 -
   1.02] respectively).
The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
prior, although the credible interval of the \textcite{niculescu2015recent}
data was narrower
  (0.07 - 
   1.34).
When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
low estimated $\alpha$ values
  (0.31 
  [0 - 
  0.73]
and
  0.38 
  [0.01 -
   0.8]). 
However, the MAP estimates increased when $m = 1$ was not permitted, although
the HPD intervals remained roughly the same
  (0.55 
  [0.01 - 
  0.93]
and
  0.5 
  [0.07 -
   0.94]).
The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
for both priors on $m$
  (0.5 for $m \geq 1$;
   0.34 for $m \geq 2$).
However, the confidence interval was much wider when $m = 1$ was allowed
  ([0.03 -
    1.61] for $m \geq 1$ vs.
   [0 -
    0.73] for $m \geq 2$).

For all the datasets except \citeauthor{novitsky2014impact}, estimated values
of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
intervals compared to the nonzero prior density region
  (\citeauthor{cuevas2009hiv}, 943 
  [287 -
   2093];
   \citeauthor{niculescu2015recent}, 1032
  [136 - 
   2851];
  \citeauthor{li2015hiv}, 1437 
  [365 -
   2897];
   \citeauthor{wang2015targeting}, 872
  [173 - 
   2119]).
The \citeauthor{novitsky2014impact} data was the outlier, with a very high
estimated $I$, and HPD interval spanning almost the entire prior region
  (5586 
  [187 -
   8806]).
The $I$ estimates and HPD intervals were generally robust to the choice of
prior on $m$, with slightly narrower HPD intervals (compare
\cref{fig:abchpd,fig:abchpdm2}).

The MAP estimate of $m$ was equal to 1 for all but the
\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
upper bound of the HPD interval was different for each dataset
  (\citeauthor{niculescu2015recent}, 5;
   \citeauthor{wang2015targeting}, 4;
   \citeauthor{li2015hiv}, 1;
   \citeauthor{cuevas2009hiv}, 2).
When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
HPD intervals spanning the entire prior region. The estimates for the total
number of nodes $N$ were largely uninformative for all samples, with almost all
MAP estimates greater than 7500 and HPD intervals spanning almost the entire
nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
data, for which the MAP estimate was lower 
  (6049)
when $m = 1$ was allowed.

@@

To make our analyses comparable to the existing network literature, we
estimated values of the power law exponent $\gamma$ for each of the datasets 
investigated. The results are summarized in \cref{tab:gamma}. All of the
estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.

\begin{table}
    \centering
    \input{\tablepath/realdata-gamma.tex}
    \caption[
        Estimated power law exponents for six HIV datasets based on maximum
        \textit{a priori} estimates of BA model parameters.
    ]{
        Estimated power law exponents for six HIV datasets based on maximum
        \textit{a priori} estimates of BA model parameters. 100 networks were
        simulated using \textit{MAP} parameter estimates obtained with
        \software{netabc}. The power law exponent $\gamma$ was estimated for
        each, and the median of those estimates was used as a point estimate
        for the corresponding dataset.
    }
    \label{tab:gamma}
\end{table}

>>>>

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. A
sudden precipitous drop in \gls{ESS} would indicate that only a small number of
particles had non-zero importance weights, which would result in a very poor
Monte Carlo approximation to the target distribution. This situation is
referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
confused with the \gls{BA} model parameter)}} controls the decay rate.
{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive},
\uline{the parameter is called $\alpha$, but to avoid confusion with the
\gls{BA} parameter of the same name, we will refer to it here as
\gls{alphaess}.}}
The tolerance $\varepsilon_i$ is chosen to satisfy
\[
  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
\]
where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
proposal distributions
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
\]
where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
simulated according to $\theta'$. In our implementation, $q$ is either a
Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
parameters. For the Poisson proposals, the number of steps to move the particle
is drawn from a Poisson distribution, and the direction in which to move the
particle is chosen uniformly at random. For both proposals, the variance was
set equal to twice the empirical variance of the particles,
following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
kernels are
\[
  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
\]
When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
with the weight update 
\begin{align*}
  w_i(x) 
    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
\end{align*}
In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
simulated datasets that are still inside the new threshold. 
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel, and the algorithm will be stopped when either of
these termination conditions is reached. The latter condition stops the
algorithm when the particles are not moving around very much, implying little
change in the estimated target.

@@

\uline{We have shown that both the normalizing constant $\Pr(T)$ and the
likelihood $\Pr(T \mid \theta)$ are likely computationally prohibitive to
calculate. If this is the case, the problem of fitting contact network models
to phylogenies seems to be of the \emph{doubly-intractible}
ype~\autocite{murray2012mcmc}, which would imply that these models are not
amenable to neither \gls{ML} nor Bayesian inference techniques. Although both
methods are able to cope with an intractible normalizing constant (for example,
by local search for \gls{ML} or Bayesian \gls{MCMC}), neither can avoid the
intractable likelihood calculations. This justifies the use of \gls{ABC}, which
is a likelihood-free method.}

@@

%{\color{blue}\uline{The \gls{BA} contact network model} (see \cref{subsec:pa})
%\uline{generates networks with a power law degree distribution. Power law
%distributions have been observed by several empirical studies of sexual
%behaviour}~\autocite{colgate1989risk, liljeros2001web, schneeberger2004scale,
%clemencon2015statistical} (although not all, see
%\autocite{handcock2004likelihood, bansal2007individual}).}

@@

{\color{blue}\uline{
Our first computational experiment was designed as an exploratory analysis of
the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
\gls{m}, and \gls{N}. The objective of this experiment was to determine whether
any of the four parameters were identifiable from the shape of the transmission
tree, as quantified by the tree kernel. Each of the \gls{BA} model parameters
was varied one at a time, while holding the other parameters at fixed, known
values. Contact networks were generated according to each set of parameter
values, and transmission trees were simulated over the networks. We then
evaluated how well a classifier based on the tree kernel could differentiate
the trees simulated under distinct parameter values. If the classifier's
cross-validation accuracy was high, this could be taken as an indication that
the parameter in question was identifiable in the range of values considered. 
A caveat of this preliminary analysis is that, since all parameters but one
were held at known values, nothing could be said about the identifiability of
\emph{combinations} of parameters; this issue will be explored later by jointly
estimating all parameters with \gls{ABC}}.

\uline{In addition to testing for identifiability, a secondary objective of
this analysis was to validate the use of the tree kernel as a distance measure
for \gls{ABC} in our context. As discussed in} \cref{sec:abc}, \uline{the
choice of distance function is extremely important for the accuracy of the
\gls{ABC} approximation to the posterior. Therefore, we evaluated two
additional tree statistics in the same manner as we evaluated the tree kernel
(that is, by constructing and testing a classifier). First, we considered
Sackin's index~\autocite{shao1990tree}, which measures the degree of imbalance
or asymmetry in a phylogeny} (see \cref{subsec:treeshape}). \uline{Sackin's
index is widely used for characterizing
phylogenies~\autocite{frost2013modelling} and has been demonstrated to vary
between transmission trees simulated under different contact network
types~\autocite{leventhal2012inferring}. Sackin's index does not take branch
lengths into account, considering only the tree's topology. The other statistic
we considered was the \gls{nltt}~\autocite{janzen2015approximate}, which
compares two trees based on normalized distributions of their branching times}
(see \cref{subsec:treeshape}).  \uline{In contrast with Sackin's index, the
\gls{nltt} does not explicity consider the trees' topologies, but it does use
their normalized branch lengths. While the \gls{nltt} is a newly developed
statistic not yet in widespread use, the unnormalized
\gls{ltt}~\autocite{nee1992tempo} was the basis of seminal early work
extracting epidemiological information from
phylodynamics~\autocite{holmes1995revealing}. We expected the tree kernel to
classify the \gls{BA} parameters more accurately than either Sackin's index or
the \gls{nltt}, since the tree kernel takes both topology and branch lengths
into account. }}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{gridsearch-example}
  \caption[
    Grid search estimates of \gls{BA} model parameters for one replicate
    experiment with trees of size 500.
  ]{
    Grid search estimates of \gls{BA} model parameters for one replicate
    experiment with trees of size 500. Point estimates (dots), 95\% highest
    density intervals (lines), and 50\% highest density intervals (notches) for
    each \gls{BA} model parameter, obtained using grid search. Networks and
    transmission trees were simulated over a grid of values for each parameter
    while holding the others fixed. For a subset of the grid values ($x$-axis),
    test networks and trees were created and compared to each tree on the grid
    using the tree kernel. The kernel scores along the grid were normalized to
    resemble a probability distribution, from which the mode and highest
    density interval were calculated. 
  } 
  \label{fig:gridest}
\end{figure}

@@

{\color{blue}
\uline{Across all simulations, the median [\gls{IQR}] absolute errors of the parameter 
estimates obtained with \software{netabc} were
    0.11 
    [0.03 - 
    0.25]
for \gls{alpha},
    492 
    [294 - 
    782]
for \gls{I},
    1 
    [0 - 
    1]
for \gls{m}, and
    4153 
    [3660 - 
    4489]
for \gls{N}. These errors comprised, respectively,
    6\%,
    11\%,
    17\%, and
    29\%
of the regions of nonzero prior density. Average \gls{HPD} interval widths were
    0.68,
    2454,
    3.01, and
    12046,
representing 
    34\%,
    55\%,
    50\%, and
    83\%
of the nonzero prior density regions. Point estimates of \gls{I} were upwardly
biased: \gls{I} was overestimated in
    69
out of 
    72
simulations
    (96\%).
The estimates for \gls{m} and \gls{N} were similar across all simulations
(median [IQR] point estimates
    3
    [3 - 
    3]
and 
    9153
    [8660 - 
    9489])
regardless of the true values of any of the \gls{BA} parameters.}

@@

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Mon Jun  6 15:27:33 2016
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
Parameter & Estimate & Standard error & $p$-value \\ 
  \hline
(Intercept) & $2$ & $0.6$ & $0.01$ \\ 
  $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
  I & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
  m & $0.5$ & $0.2$ & $0.01$ \\ 
   \hline
\end{tabular}
\caption[Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters. GLM was fitted with a Gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glmalpha}
\end{table}
% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Mon Jun  6 15:27:33 2016
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
Parameter & Estimate & Standard error & $p$-value \\ 
  \hline
(Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
  $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
  I & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
  m & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
   \hline
\end{tabular}
\caption[Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters. GLM was fitted with a gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glmI}
\end{table}

@@

@@

Empirical studies of contact networks often report the exponent $\gamma$ of the
power law degree distribution. To compare our results to the literature, we
simulated 100 networks each according to the posterior mean parameter estimates
obtained for each investigated dataset. The power-law exponent $\gamma$ was
calculated for each network using the \software{fit\_power\_law} function in
\software{igraph}, with the `R.mle' implementation. The median of the 100
$\gamma$ values was taken as a point estimate for the associated dataset.

@@

We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
and found substantial heterogeneity among the parameter estimates
(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
for each dataset are shown in
\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
mean [95\% \gls{HPD}] 
  0.73 
  [0.05 - 
   1.18]
and
  0.55 
  [0.01 -
   0.99] respectively).
The estimates did not change appreciably when $m = 1$ was disallowed by the
prior, although the credible interval of the \textcite{niculescu2015recent}
data was narrower
  (0.05 - 
   1.18).
When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
low estimated $\alpha$ values
  (0.33 
  [0 - 
  0.76]
and
  0.27 
  [0 -
   0.59]). 
However, the estimates increased when $m = 1$ was not permitted, although the
HPD intervals remained roughly the same
  (0.58 
  [0.06 - 
  0.99]
and
  0.48 
  [0.02 -
   0.87]).
The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
for both priors on $m$
  (0.55 for $m \geq 1$;
   0.53 for $m \geq 2$).
However, the confidence interval was much wider when $m = 1$ was allowed
  ([0 -
    1.75] for $m \geq 1$ vs.
   [0 -
    1.75] for $m \geq 2$).

For all the datasets except \citeauthor{novitsky2014impact}, estimated values
of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
intervals compared to the nonzero prior density region
  (\citeauthor{cuevas2009hiv}, 701 
  [289 -
   1279];
   \citeauthor{niculescu2015recent}, 747
  [136 - 
   2378];
  \citeauthor{li2015hiv}, 1390 
  [310 -
   2821];
   \citeauthor{wang2015targeting}, 675
  [175 - 
   1400]).
The \citeauthor{novitsky2014impact} data was the outlier, with a very high
estimated $I$, and HPD interval spanning almost the entire prior region
  (5431 
  [183 -
   8739]).
The $I$ estimates and HPD intervals were generally robust to the choice of
prior on $m$, with slightly narrower HPD intervals (compare
\cref{fig:abchpd,fig:abchpdm2}).

The posterior mean of $m$ was equal to 1 for all but the
\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
upper bound of the HPD interval was different for each dataset
  (\citeauthor{niculescu2015recent}, 4;
   \citeauthor{wang2015targeting}, 1;
   \citeauthor{li2015hiv}, 1;
   \citeauthor{cuevas2009hiv}, 1).
When $m = 1$ was disallowed, the estimate for all datasets was either 2 or 3,
with HPD intervals spanning the entire prior region. The estimates for the
total number of nodes $N$ were largely uninformative for all samples, with
almost all posterior mean estimates greater than 7500 and HPD intervals
spanning almost the entire nonzero prior density region. The only exception was
the \citeauthor{li2015hiv} data, for which the estimate was lower 
  (5916)
when $m = 1$ was allowed.

@@

To make our analyses comparable to the existing network literature, we
estimated values of the power law exponent $\gamma$ for each of the datasets 
investigated, using the posterior median estimates with the prior allowing
\gls{m} = 1. The results are summarized in \cref{tab:gamma}. All of the
estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.

\begin{table}
    \centering
    \input{\tablepath/realdata-gamma.tex}
    \caption[
        Estimated power law exponents for six HIV datasets based on maximum
        \textit{a priori} estimates of BA model parameters.
    ]{
        Estimated power law exponents for six HIV datasets based on maximum
        \textit{a priori} estimates of BA model parameters. 100 networks were
        simulated using posterior mean parameter estimates obtained with
        \software{netabc}. The power law exponent $\gamma$ was estimated for
        each, and the median of those estimates was used as a point estimate
        for the corresponding dataset.
    }
    \label{tab:gamma}
\end{table}
